---
title: 'Formative Assessment 2: Extras'
author: "John Benedict A. Monfero"
date: "2024-02-12"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# install.packages("latex2exp")
# library(latex2exp)
```
# Number Two
## An experiment consists of tossing two fair coins.
### Use R to simulate this experiment 100 times and obtain the relative frequency of each possible outcome. Hence, estimate the probability of getting one head and one tail in any order.
```{r}
# Enumerate the sample space whenever two fairs coins tossed:
sample_space <- c("HH", "HT", "TH", "TT")
# In R, we can simulate the relative frequency of the tossing two coin experiment with 100 repetitions through sample() function
experiment_100 <- sample(sample_space, 100, replace = T)

# Preliminary Program Output
print("Here is the frequency distribution of the experiment of tossing two fair coins")
print("HH, and TT are the not desired events, HT and TH are favorable events")
table(experiment_100)

par(bg = "#f3f3f3")
HH_events <- sum(experiment_100 == "HH")
favored_events <- sum(experiment_100 == "HT" | experiment_100 == "TH")
TT_events <- sum(experiment_100 == "TT")

barplot(c("HH" = HH_events, "HT + TH" = favored_events, "TT" = TT_events),  main = "Experiment consists two tossing coins", xlab = "Events in tossing two coins", ylab = "Total Frequency", col = c("steelblue", "gold", "steelblue"), las = 1)
```


$$
P(HT\hspace{3 pt} \cup \hspace{3 pt} TH) = P(HT) + P(TH) - P(HT\hspace{3 pt}  \cap \hspace{3 pt} TH)
$$
```{r}
# Calculate the total number of events
total_events <- HH_events + favored_events + TT_events

# Calculate the relative frequency for each outcome
relative_frequency_HH <- HH_events / total_events
relative_frequency_favored <- favored_events / total_events
relative_frequency_TT <- TT_events / total_events

# Output the relative frequencies
relative_frequencies <- c("HH" = relative_frequency_HH, "HT or TH" = relative_frequency_favored, "TT" = relative_frequency_TT)
print(paste("Hence the estimation of probability of getting a head and a tail in any order would be", relative_frequency_favored*100, "%"))
relative_frequencies
```
# Number Three
##  An experiment consists of rolling a die. 
### Use R to simulate this experiment 600 times and obtain the relative frequency of each possible outcome. Hence, estimate the probability of getting each of 1, 2, 3, 4, 5, and 6.

```{r}
# When a die (one dice) rolled, enumerate the event possibilities that an even number appears.
even <- c(2,4,6)
# When a die (one dice) rolled, enumerate the event possibilities that not an even number appears.
odd <- c(1,3,5)

# A die whose holding its universal events (sample).
die <- c(even, odd)

# Store the result of the experiment 600 times in a variable
roll_results <- sample(die, 600, replace = T)

# Count the number of times each result appears
roll_counts <- table(roll_results)

par(bg = "#f3f3f3")
# Create a bar plot with these counts
barplot(roll_counts, main = "Summed Frequencies", xlab = "Die Face Values", ylab = "Total Frequency", 
        col = c("salmon", "lightgreen", "salmon", "lightgreen", "salmon", "lightgreen"), las = 1)
```

### Since the experiment is made 600 times, therefore the sample space also has 600 events where all came from 6 different favorable events
$$
\text{Sample Space} \hspace{3pt}= \hspace{3pt} 600 \hspace{3pt} \text{Events} \hspace{3pt} = \hspace{3pt} E_1 + E_2 + E_3 + E_4 + E_5 + E_6
$$

```{r}
# Determine the relative frequencies on each events
roll_counts/600

print(paste("To begin, the relative probability on getting 1 on the given rolling a die experiment is: ", round(roll_counts[1]/600, 4)))
print(paste("Secondly, the relative probability on getting 2 on the given rolling a die experiment is: ", round(roll_counts[2]/600, 4)))
print(paste("Moreover, the relative probability on getting 3 on the given rolling a die experiment is: ", round(roll_counts[3]/600, 4)))
print(paste("Fourthly, the relative probability on getting 4 on the given rolling a die experiment is: ", round(roll_counts[4]/600, 4)))
print(paste("Subsequently, the relative probability on getting 5 on the given rolling a die experiment is: ", round(roll_counts[5]/600, 4)))
print(paste("Lastly, the relative probability on getting 6 on the given rolling a die experiment is: ", round(roll_counts[6]/600, 4)))
```

### The sample probability shall be 100%
$$
P(S)= 1 = P(E_1)+P(E_2)+P(E_3)+P(E_4)+P(E_5)+P(E_6)
$$

```{r}
prob_sample_space_die <- roll_counts[1]/600 + roll_counts[2]/600 + roll_counts[3]/600 + roll_counts[4]/600 + roll_counts[5]/600 + roll_counts[6]/600
print(paste("The probability of the sample space {getting 1,2,3,4,5 or 6} from a rollng die would be: ", prob_sample_space_die*100,"%"))
```

# Number Four
## Amy and Jane are gambling together
### A fair coin is tossed repeatedly. Each time a head comes up, Amy wins two euro from Jane, and each time a tail comes up, Amy loses two euro to Jane. Use R to simulate this game 100 times, and estimate:

```{r}
# Function to calculate the cumulative sum of winnings or losses
calculate_cumulative_total <- function(gambling_simulation) {
  heads_count <- cumsum(gambling_simulation == 2)
  tails_count <- cumsum(gambling_simulation == -2)
  # Assuming Amy wins on heads and Jane wins on tails
  amy_total <- heads_count - tails_count
  return(amy_total)
}

# Run one long simulation
gambling_simulation <- sample(c(2, -2), 100, replace = TRUE)

# Calculate the cumulative total after each trial
gambling_result <- calculate_cumulative_total(gambling_simulation)

par(bg = "#f3f3f3")
# Plot the progression of cumulative winnings or losses
plot(1:100, gambling_result, type = "o", col = "blue",
     xlab = "Number of Tosses", ylab = "Cumulative Total", main = "Amy's Progression of Winnings/Losses")

# Optional: Add a horizontal line at y=0 to show the break-even point
abline(h = 0, col = "red", lty = 2)

# Highlight the last trial
last_trial <- length(gambling_result)
last_value <- gambling_result[last_trial]

# Add a point for the last trial
points(last_trial, last_value, col = "darkgreen", pch = 19, cex = 1.5)

# Annotate the plot with the value of the last trial
text(last_trial, last_value, labels = paste(last_value, sep=""), cex = 0.5, col = "darkgreen", pos = 4)

```


#### Estimate the number of times that Amy is ahead in these 100 tosses;
```{r}
# Calculate the number of times Amy is ahead
times_ahead <- sum(gambling_result > 0)

# Print the number of times Amy is ahead
print(paste("In the gambling simulation so far, Amy is winning ahead from debt or break-even line by", times_ahead, "times!"))
```

#### Estimate how much Amy has won or lost after 100 tosses.
```{r}
# Assuming gambling_result contains the cumulative total of Amy's winnings/losses after 100 tosses
final_result <- gambling_result[100]

# Conditional statements to determine the message output
if (final_result > 0) {
  message <- paste("After 100 tosses, Amy is ahead by", final_result, "euros.")
} else if (final_result < 0) {
  message <- paste("After 100 tosses, Amy is behind by", abs(final_result), "euros.")
} else {
  message <- "After 100 tosses, Amy has broken even."
}

# Print the message
print(message)
```

